Probabilistic Catalogs for Crowded Stellar Fields

### 2012-10-26 paper/ms.tex
SExtractor results figure
Some writing tweaks

\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{bj.brewer@auckland.ac.nz}

@@

\begin{abstract}We introduce a probabilistic (Bayesian) method for producing catalogs
from images of crowded stellar fields. The method is capable of handling the
challenges introduced by overlapping sources and is able to determine the luminosity
function of the stars even when the precise luminosity of each star is uncertain.
This is in contrast with standard techniques which make a single catalog,
potentially underestimating the uncertainties in any study of the stellar
population. The method is implemented using advanced Markov Chain Monte Carlo
(MCMC) techniques including Reversible Jump and Nested Sampling.
The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. The parameters of the luminosity function can be recovered
with moderate uncertainty. We compare the results obtained from our method
with those obtained from the SExtractor software and find that the latter is
unable to correctly infer the number of stars or the luminosity function
parameters.
\end{abstract}

@@

\section{Introduction}
A fundamental problem in astronomy is the construction of {\it catalogs} from
raw image data. {\bf The catalogs are then used for...}
Standard tools for generating catalogs include \texttt{SExtractor}
\citep{sextractor}, {\bf more refs}.
\citep{2011EAS....45..351H}

@@

\section{Bayesian Inference}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

@@

To proceed with the model for inferring catalogs from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}.

\section{The Specific Model for Stellar Fields}\label{sec:model}
\subsection{The Hypothesis Space}
The hypothesis space is the set of possible catalogs, or the set of possible
answers to the question, ``What objects are present in the field and
what are their properties?'' There are an unknown number of stars
$N$ in the field. Each star has an unknown
position $(x,y)$ in the plane of the sky, and an unknown flux $f$. We also
describe the distribution of fluxes (commonly known as the {\it luminosity
function}) of the stars by some parameters denoted collectively by $\beta$.
In summary, the unknown parameters are:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.33]{Figures/summaries.eps}
\end{center}
\caption{Summary images for test case 2.\label{fig:summaries}}
\end{figure}

@@

\bibitem[\protect\citeauthoryear{Skilling}{1998}]{massinf} 
Skilling J., 1998, Massive Inference and Maximum Entropy, in Maximum Entropy 
and Bayesian Methods, Kluwer Academic Publishers, Dordrecht/Boston/London p.14
\end{thebibliography}

>>>>

\newcommand{\sex}{\texttt{SExtractor }}

\slugcomment{To be submitted to the Astronomical Journal}

@@

\begin{abstract}
We introduce a probabilistic (Bayesian) method for producing catalogs
from images of crowded stellar fields. The method is capable of handling the
challenges introduced by overlapping sources and is able to determine the luminosity
function of the stars even when the precise luminosity of each star is uncertain.
This is in contrast with standard techniques which produce a single catalog,
potentially underestimating the uncertainties in any study of the stellar
population. The method is implemented using advanced Markov Chain Monte Carlo
(MCMC) techniques including Reversible Jump and Nested Sampling.
The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. The parameters of the luminosity function can be recovered
with moderate uncertainty. We compare the results obtained from our method
with those obtained from the \sex software and find that the latter is
unable to correctly infer the number of stars or the parameters of the luminosity
function.
\end{abstract}

@@

\section{Introduction}
A fundamental problem in astronomy is the construction of {\it catalogs} from
raw image data. {\bf The catalogs are then used for...}
Standard tools for generating catalogs include \sex
\citep{sextractor}, DAOPHOT \citep{1987PASP...99..191S}.
%\citep{2011EAS....45..351H}

@@

\section{Bayesian Inference}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta).
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data sets, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

@@

To proceed with the model for inferring catalogs (stellar positions and fluxes)
from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}. Throughout this paper we make a number of simplifications
for the purposes of testing the computational feasibility and performance of
the method. In real applications, some of these assumptions (such as known PSF
and noise variance) may be relaxed.

\section{The Specific Model for Stellar Fields}\label{sec:model}
\subsection{The Hypothesis Space}
The hypothesis space is the set of possible catalogs, or the set of possible
answers to the question, ``What objects are present in the field and
what are their properties?'' We shall assume that there are an unknown number of stars
$N$ in the field. Each star has an unknown
position $(x,y)$ in the plane of the sky, and an unknown flux $f$. We also
describe the distribution of fluxes (commonly known as the {\it luminosity
function}) of the stars by some parameters denoted collectively by $\beta$.
In summary, the unknown parameters are:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{Figures/summaries.eps}
\end{center}
\caption{Summary images for test case 2.\label{fig:summaries}}
\end{figure}

\section{Comparison to SExtractor}

\begin{figure}
\begin{center}
\includegraphics[scale=0.45]{Figures/sex_N.eps}
\end{center}
\caption{The number of stars in the catalogs produced by \sex.\label{fig:sex_N}}
\end{figure}

@@

\bibitem[\protect\citeauthoryear{Skilling}{1998}]{massinf} 
Skilling J., 1998, Massive Inference and Maximum Entropy, in Maximum Entropy 
and Bayesian Methods, Kluwer Academic Publishers, Dordrecht/Boston/London p.14

\bibitem[Stetson(1987)]{1987PASP...99..191S} Stetson, P.~B.\ 1987, \pasp, 
99, 191 
\end{thebibliography}
